# Week 2: Regression and model validation 

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
library(dplyr)
```



The data used in this task is a subset of Learning 2014 survey data, collected by Kimmo Vehkalahti for an international survey of Approaches to Learning, made possible by Teachers' Academy funding for Kimmo Vehkalahti in 2013-2015.

# Reading the data, looking at structure and checking that the dataframe is readable 
```{r}
lrn2014 <- read.csv(file.path(path, "lrn2014.csv"))
str(lrn2014)
head(lrn2014)
dim(lrn2014)
```

There are 166 observations of 7 variables in the dataframe. Variables used are gender ("gender"), age ("age"), attitude ("attitude"), deep questions ("deep"), strategic questions ("stra"), surface questions ("surf") and points ("points"). Deep, stra and surf are combined variables from several related questions in the survey. Points values of 0 have been excluded from the dataframe.

# step 1 completed




