# Week 3: Logistic regression

```{r, echo=F, message=F, warning=F}
date()
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, forcats, haven, janitor, psych, lm.beta, broom, writexl,sjstats, knitr, pwr, dplyr, GGally, ggplot2)
library(dplyr)
library(GGally)
library(ggplot2)
```
This week I'm analysing a joined dataset that includes two student alcohol consumption data sets. Data set information can be found here: https://archive.ics.uci.edu/ml/datasets/Student+Performance 
The following adjustments have been made:
The variables not used for joining the two data have been combined by averaging (including the grade variables)
* 'alc_use' is the average of 'Dalc' and 'Walc'
* 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise


# What's in the dataframe
```{r}
alc <- read.csv("C:/Users/35840/OneDrive - University of Helsinki/IODS/IODS-project/data/alc.csv")
str(alc)
head(alc)
dim(alc)
```
There are 370 observations over 35 variables. Respondents are studens whose age range from 15 to 22 years, and there are binary variables such as students' sex (F for female, M for male), the school they are from, family size and so on. 
