# Week 3: Logistic regression

```{r, echo=F, message=F, warning=F}
date()
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, forcats, haven, janitor, psych, lm.beta, broom, writexl,sjstats, knitr, pwr, dplyr, GGally, ggplot2)
library(dplyr)
library(GGally)
library(ggplot2)
```
This week I'm analysing a joined dataset that includes two student alcohol consumption data sets. Data set information can be found here: https://archive.ics.uci.edu/ml/datasets/Student+Performance 
The following adjustments have been made:
The variables not used for joining the two data have been combined by averaging (including the grade variables)
* 'alc_use' is the average of 'Dalc' (weekday alcohol usage) and 'Walc' (weekend alcohol usage)
* 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise
The purpose of my analysis is to study the relationships between high/low alcohol consumption and some of the other variables in the data.

## What's in the dataframe
```{r}
alc <- read.csv("C:/Users/35840/OneDrive - University of Helsinki/IODS/IODS-project/data/alc.csv")
str(alc)
colnames(alc)
dim(alc)
```
There are 370 observations over 35 variables. Respondents are studens whose age range from 15 to 22 years, and there are numeric, nominal and  binary variables. Binary variables include questions such as students' sex (F for female, M for male), the school they are from (school), family size (famsize).

## Looking more closely at 4 variables in relation to alcohol consumption: 
I believe that it will be interesting to further study these 4 specific variables and how they relate to alcohol consumption.
* variable "school" because peer pressure can affect alcohol consumption
* variable "famsize" which is a binary variable of less than 3 family members or more than 3 family members. My hypothesis is that family size does not predict alcohol consumption.
* variable "age" because there is a legal age (let's say its 18 for this dataset) for using alcohol, and some of the respondents in the data are below 18 years old
* variable "absences", number of school absences (numeric: from 0 to 93)
 because other norm-breaking behavior such as absences from school may occur if student is also consuming alcohol regularly.
Here is an overview of the four variables in the dataset:
```{r}
alc %>% group_by(school, famsize, age, absences) %>% summarise(count = n())
```
The relationships of variables to alcohol consumption by sex

```{r}
# initialize a plot of alc_use and each variable
g1 <- ggplot(alc, aes(x = alc_use, y = school, col = sex))
g2 <- ggplot(alc, aes(x = alc_use, y = famsize, col = sex))
g3 <- ggplot(alc, aes(x = alc_use, y = age, col = sex))
g4 <- ggplot(alc, aes(x = alc_use, y = absences, col = sex))

# define the plot as a boxplot and draw it
par(mfrow = c(2,2))
g1 + geom_boxplot() + ylab("school") + ggtitle("Student school by alcohol consumption and sex")
g2 + geom_boxplot() + ylab("family size") + ggtitle("Student family size by alcohol consumption and sex")
g3 + geom_boxplot() + ylab("age") + ggtitle("Student age by alcohol consumption and sex")
g4 + geom_boxplot() + ylab("absences") + ggtitle("Student absences by alcohol consumption and sex")
```

Family size matters for female students more, but more alcohol using male students come from bigger (GT3) families whereas female students alcohol usage is lower than males' in any case.
As expected, school matters, but it seems that the effect is limited to male students, who use use alcohol more than female students. Among female students there is no difference between schools.
From the data we can see that the absences do not seem to have a specific influence on alcohol consumption.
The surprising result is that most alcohol using female students seem to be 16-17 years of age, whereas male students report using alcohol until they are 18. This relates either to the fact that the legal age of alcohol consumption is 16 (as there are only few observations of below 16 years old using alcohol) and that after the initial one or two years of legal consumption, the usage lessens significantly.

## Logistic regression analysis
```{r}
m <- glm(high_use ~ school + famsize + age + absences, data = alc, family = "binomial")
summary(m)
coef(m)
# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)

```
For one unit increase in absences, the odds of high alcohol usage increase by a factor of 1,09.There appears a statistically significant correlation between absences and high alcohol usage. 
Interestingly, all of school=MS, family size = less than 3, and age include 1 within the 95 % confidence interval, which implies that the result is not statistically significant. This result rejects my personal hypothesis as the data does not show obvious connections between the chosen variables and high alcohol usage.

## Predictive power of the model
The only variable with a statistically significant connection to high alcohol usage is "absences". I'll explore its prediction capabilities next. Let's look at the cross tabulation of the prediction versus actual values of high usage:
```{r}
# fit the model
m2 <- glm(high_use ~ absences, data = alc, family = "binomial")

# predict() the probability of high_use
probabilities <- predict(m2, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = (probability > 0.5))

# see the last ten original classes, predicted probabilities, and class predictions
select(alc, absences, high_use, probability, prediction) %>% tail(10)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction)
```

```{r}
# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use, col = prediction))

# define the geom as points and draw the plot
g + geom_point()

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins
```
Based on this, the prediction is surprisingly good at predicting lack of high usage (false values accurate 68,6 % of the time).

```{r}

# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)
```