# Week 5: Dimensionality reduction techniques

# Making libraries available
```{r dpi = 200}
date()
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, forcats, haven, janitor, psych, lm.beta, broom, writexl,sjstats, knitr, pwr, corrplot, tidyr, dplyr, ggplot2, GGally, knitr, png)
library(knitr)
library(png)
library(corrplot)
library(tidyr)
library(dplyr)
library(ggplot2)
library(GGally)
```

This week we are learning about dimensionality reduction technique. The dataset in use is a combined and modified "human" dataset, which originates from United Nations Development Programme data of Human development index and Gender Equality. Here you can find more information of the data: http://hdr.undp.org/en/content/human-development-index-hdi

The variables are 
"Life.Exp" for Life expectancy at birth,
"Edu.Exp" for Expected years of education,
"GNI" for Gross National Income per capita,
"Mat.Mor" for Maternal mortality ratio,
"Ado.Birth" for Adolescent birth rate, 
"Parli.F" for Percent of Representation in Parliament, 
"Edu2.FM" for the ratio of Female and Male populations with secondary education in each country
"Labo.FM" for the  ratio of Female and Male populations labour force participation in each country. 
```{r}
# load the data from my computer and telling R that the row names are in the first column
human <- read.csv("C:/Users/35840/OneDrive - University of Helsinki/IODS/IODS-project/data/human.csv", row.names = 1)

# explore the dataset
str(human)
dim(human)
summary(human)
```

There are 155 observations of 9 variables, with all other variables as numeric except Mat.mor. 
A graphical overview with histograms, and showing correlations between variables:
```{r}
# histograms
gather(human) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_histogram()
# calculate the correlation matrix and round it
cor_matrix<-cor(human) %>% round(digits = 2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6,
         col = COL2('PRGn'))
```

It looks like there is very little correlation between the variables Parli.F and Labo.FM with the other variables. Most understandably there is strong correlation between Life.Exp and Edu.Exp, as we can assume that both life expectancy and years of secondary education are affected by general living standards and well-being. Accordingly, a strong negative correlation is shown between Mat.Mor and Life.Exp. GNI's correlation with the variables is in general surprisingly weak, when we think that Gross National Income would play a role in the general living standards.

Next, we'll perform a principal component analysis (PCA) to the non-standardized data:
```{r}
# principal component analysis (with the SVD method) and the variability captured by the principal components
pca_human <- prcomp(human)
summary(pca_human)
```
The variation captured is noted on the line "Proportion of variance" and we can see that PC1 captures almost 77% of the variation and PC2 about 22%, which leaves only about 1% of the variation causes to the other components.

Let's look at a biplot of the PC representation with the first two variables:
```{r, out.width="300%", fig.cap="Non-Standardized dataset. Variables affecting the Human Development Index"}
# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))

```

Now we'll try that again but with standardized data:
```{r}
# standardize the variables
human_std <- scale(human)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis (with the SVD method) and check variation
pca_human2 <- prcomp(human_std)
summary(pca_human2)
```

```{r, out.width="300%", fig.cap="Standardized dataset. Variables affecting the Human Development Index"}
# draw a biplot of the principal component representation and the original variables
biplot(pca_human2, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

Now the variance captured by PC1 is only around 50% and PC2 around 16%, which evens out the variation caused by other components, while we are still able to say that the variation is mainly caused by the two first components. 
The results are different with standardized and non-standardized data. 
The non-standardized data mapped on the biplot only show the first two principal components, as they together explain almost all of the variance. Sub-saharan African countries are singled out. The standardized data analyzed with PCA on the other hand shows that more factors explain the variance. Northern Europe stands out as a cluster, explained by Parliamentary representation, ratio of F/M labour and expected years of secondary education. 

