# Week 6: Analysis of longitudinal data

Making libraries available
```{r}
date()
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, forcats, haven, janitor, psych, lm.beta, broom, writexl,sjstats, knitr, pwr, corrplot, tidyr, dplyr, ggplot2, GGally,lme4)
library(corrplot)
library(tidyr)
library(dplyr)
library(ggplot2)
library(GGally)
library(lme4)
```

This week we are analysing longitudinal data with not-related datasets about weights of rats (RATS) and symptom assessment on human males (BPRS).

After reading the datasets to R, we'll have to convert the first two variables to factors. For visualization purposes, new columns to help see the passed time will be created: "week" to BPRS and "Time" to RATS.

```{r}
# BPRS data:
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# RATS data:
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep  ="\t", header = T)
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Converting both datasets to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
RATSL <- RATS %>% gather(key = WD, value = Weight, -ID, -Group)

# Add the week column to BPRSL
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))

# Add the time column to RATSL
RATSL <- RATSL %>% mutate(Time = as.integer(substr(WD,3,4))) 


# Glimpse the data to check variable names and data contents
glimpse(RATSL)
glimpse(BPRSL)
```

## RATS data

Then we're going to begin the analyses, first with RATS (longform) data:
```{r}
# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```
The weights of the rats in the group 1 are consistently lower than in groups 2 and 3, where there are heavier rats.
Now we take a new go with standardised data:
```{r}

# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)
str(RATSL)
# Plot again with the standardised bprs
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardised weight")

```

Then we'll create a summary and plot the mean profiles:
```{r}
# Number of weeks, baseline (week 0) included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Now we'll create a summary data by Group and ID with mean as the summary variable (ignoring baseline point in Time at 1 days)
```{r}
RATS8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time (in days) 1-64")
```

There's a curious outlier, a too heavy rat in group 2, which we'll remove and draw the plot again:
```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATS8S1 <- RATS8S %>%
  filter(mean < 550)

ggplot(RATS8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time (in days) 1-64")
```

Finally, we'll add the first poin in time as baseline (Wd1) and create an analysis of variance (anova) table for the RATS:
```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATS8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

## BPRS data

Now we'll analyse the BPRS data:
```{r}
# Check the dimensions and insides of the data
dim(BPRSL)
str(BPRSL)

# Plot the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

``` 

Next up, we'll fit a regression model to the data
``` {r}

# create a regression model RATS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

Then we'll create a random intercept model:
```{r}
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```
To allow for closer inspection of individuals, let's fit the random intercept and random slope model to the BPRS data: 
```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```
Finally, we'll draw a couple of plots:
```{r}
# create a random intercept and random slope model
BPRS_ref2 <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of RATSL
ggplot(BPRSL, aes(x = week, y = bprs, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "Observed symptoms (bprs)") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(BPRSL, aes(x = week, y = Fitted, group = treatment)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "Fitted symptom values (bprs)") +
  theme(legend.position = "top")


```
The overall interpretation is that over time, the symptoms are reduced with treatment, even though there is a lot of variation between treatment types.