# Week 5: Dimensionality reduction techniques

# Making libraries available
```{r}
date()
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, forcats, haven, janitor, psych, lm.beta, broom, writexl,sjstats, knitr, pwr, corrplot, tidyr, dplyr, ggplot2, GGally)
library(MASS)
library(corrplot)
library(tidyr)
library(dplyr)
library(ggplot2)
library(GGally)
```

This week we are learning about dimensionality reduction technique. The dataset in use is a combined and modified "human" dataset, which originates from United Nations Development Programme data of Human development index and Gender Equality. Here you can find more information of the data: http://hdr.undp.org/en/content/human-development-index-hdi

```{r}
# load the data from my computer and telling R that the row names are in the first column
human <- read.csv("C:/Users/35840/OneDrive - University of Helsinki/IODS/IODS-project/data/human.csv", row.names = 1)

# explore the dataset
str(human)
dim(human)
summary(human)
```

There are 155 observations of 9 variables, with all other variables as numeric except Mat.mor. 